---
title: "Lab8"
author: "Brandon Nelsen, Tanner Ball, Burke Parent, Drew Hoffman, David Washburn"
date: "10/14/2019"
output: html_document
---
```{r}
library(tidyverse)
library(readxl)
```

***Team Portion***

1. Who is hiring you?

answer here

2. Overall Team Question, why it's important, and why those reasons are important

answer here

3. A brief description of the dataset you want to use

answer here

4. A series of individual questions and why they're important

answer here

5. Each team member should produce a different visualization of the East Africa Lake Data. Create plots that show the data over time for Challa, Tanganyika, Malawi, CRU and MAM rainfall, and VOI rainfall.

```{r}
# Challa Data Set - Tanner
Challa <- read_xlsx("East Africa lake data.xlsx", sheet = "Challa")
Challa <- filter(Challa, !is.na(`Age AD`))
Challa <- transmute(Challa,year=`Age AD`, number=`varve number`, thickness=as.double(`varve thickness`),dark.layer=as.double(`dark layer`), light.layer=as.double(`light layer`))
Challa
```

```{r}
#Tanganyika Data Set
Tanganyika <- read_xlsx("East Africa lake data.xlsx", sheet = "Tanganyika")
Tanganyika$X3 = NULL
Tanganyika$X6 = NULL
Tanganyika$X7 = NULL 
Tanganyika$X10 = NULL
Tanganyika$X11 = NULL
Tanganyika$X12 = NULL

TidyTanganyika <- Tanganyika
TidyTanganyika %>% unite('Year', 'Year AD')
```
```{r}
#naivasha Data Set - David
Naivasha <- read_xlsx("East Africa Lake Data.xlsx", sheet=5)
    Year = col_date(format="%Y")
Naivasha$Year <- round(Naivasha$Year)
Naivasha <- rename(Naivasha,LakeDepth=`Lake Depth`)

ggplot(Naivasha, aes(Year, LakeDepth)) +
  geom_point()
``` 
